Fitting the undetected: Light-curve feature estimation in the presence of carelessly reported non-detections

### 2011-11-14 tex/ms.tex
Added manuscript TeX, figure, ASAS data files, and simulated data files
git-svn-id: svn+ssh://astrometry.net/svn/trunk/projects/censored@19907 03a3cea6-2c03-0410-ac9b-9271b1c66c29

% Copyright 2011 David W. Hogg (NYU).
% All rights reserved.

\documentclass[12pt]{article}

@@

\begin{document}\sloppy\sloppypar\raggedbottom

\textsl{This document is a draft dated 2011-11-09.  It incorporates
  direct contributions from Joseph~W.~Richards (Berkeley),
  David~W.~Hogg (NYU), James~P.~Long (Berkeley), and
  Daniel~Foreman-Mackey (NYU) and indirect contributions from many
  others.}

\vspace{1ex}

@@

@@

Under the assumption that the inclusion (or not) of the source in the
catalog is based on (or very strongly related to) the measured flux
(rather than the true flux), the natural likelihood for the bit $q_i$,
on which the above likelihoods are conditioned is
\begin{eqnarray}\displaystyle
P(q_i|b_i,\sigma_i,\theta,I) &=& \left\{\begin{array}{ll}
  \displaystyle\int_0^{b_i} N(f|\mu_i,s_\mu^2+\sigma_i^2)\,\dd f & \mbox{for $q_i=0$} \\
  \displaystyle\int_{b_i}^\infty N(f|\mu_i,s_\mu^2+\sigma_i^2)\,\dd f & \mbox{for $q_i=1$}
\end{array}\right.\label{eq:pofq}
\quad,
\end{eqnarray}
where the integrals end at an observed-flux limit $b_i$, which (as
mentioned above) is also a model parameter.  The motivation for this
expression is that it imagines all possible observations $f$ of the
star at time $t_i$ given the model and uncertainties, and finds the
fraction that would make the (imagined) detection cut.

@@

>>>>

% Define the document class. aastex for manuscript preparation, emulateapj for making it look polished.
% \documentclass[12pt,preprint2]{aastex}
%\documentclass[12pt,a4paper, iop,onecolumn,numberedappendix]{emulateapj}
\documentclass[12pt,preprint]{aastex}

% Import the natbib package for bibliography inclusion. To get it to work right, first typeset the bibliography (BibTex), Command-Shift-B.
% Then, typeset the document with LaTeX, Command-Shift-L or just Command-T.
% To get proper reference citations using aastex (preprint manuscripts), first typeset as emulateapj with BibTex to create the BibTex 
% files and then typeset again as aastex to get the manuscript with references.
\usepackage{natbib}
\usepackage{epsfig,graphicx,color}
\usepackage{longtable}

\bibliographystyle{apj}
% A few colors to replace the defaults for certain link types
\definecolor{orange}{cmyk}{0,0.4,0.8,0.2}
\definecolor{darkorange}{rgb}{.71,0.21,0.01}
\definecolor{darkgreen}{rgb}{.12,.54,.11}

%-----------------------------------------------------------------------------
% The hyperref package gives us a pdf with properly built
% internal navigation ('pdf bookmarks' for the table of contents,
% internal cross-reference links, web links for URLs, etc.)
\usepackage{hyperref}
\hypersetup{pdftex,  % needed for pdflatex
  breaklinks=true,  % so long urls are correctly broken across lines
  colorlinks=true,
  urlcolor=blue,
  linkcolor=darkorange,
  citecolor=darkgreen,
  }

%Define this shorthand for scientific notation.
\newcommand{\expnt}[2]{\ensuremath{#1 \times 10^{#2}}}   % scientific notation
\newcommand{\x}{{\bf x}}
\newcommand{\X}{{\bf X}}
\newcommand{\y}{{\bf y}}
\newcommand{\z}{{\bf z}}
\newcommand{\new}{red}
\newcommand{\ptr}{p_{\rm tr}}
\newcommand{\pte}{p_{\rm te}}
\newcommand{\Ex}{\mathbf{E}}
\newcommand{\Prob}{\mathbf{P}}
\newcommand{\PrfL}{\widehat{P}_{\rm{RF,}\mathcal{L}}(y|\x)}
\newcommand{\PrfLx}{\widehat{P}_{\rm{RF,}\mathcal{L}\cup\x'}(y|\x)}
\newcommand{\TreeL}{\theta_{b, \mathcal{L}}(y|\x)}
\newcommand{\TreeLx}{\theta_{b, \mathcal{L}\cup\x'}(y|\x)}
 \def\hipp {{\it Hipparcos~}}

@@

% Commands for annotating the docs with fixme and inter-author notes.  See
% below for how to disable these.
%
% Define a \fixme command to mark visually things needing fixing in the draft,
% as well as similar commands for each author to leave initialed special
% comments in the document.
% For final printing or to simply disable these bright warnings, symlink
% (there's a target 'dist_on' in the makefile that does this) the file
% macros_state.tex to macros_off.tex

\newcommand{\fixme}[1] { \textcolor{red} {
{\fbox{ {\bf FIX}
\ensuremath{\blacktriangleright \blacktriangleright \blacktriangleright}}
{\bf #1}
\fbox{\ensuremath{\blacktriangleleft \blacktriangleleft \blacktriangleleft}
} } } }

% And similarly, one (less jarring, with fewer symbols and no boldface) command
% for each one of us to leave comments in the main text.
\newcommand{\james}[1] { \textcolor{blue} {
\ensuremath{\blacklozenge} {\bf james:}  {#1}
\ensuremath{\blacklozenge} } }

\newcommand{\hogg}[1] { \textcolor{darkorange} {
\ensuremath{\blacksquare} {\bf hogg:}  {#1}
\ensuremath{\blacksquare} } }

\newcommand{\dan}[1] { \textcolor{darkgreen} {
\ensuremath{\bigstar} {\bf dan:}  {#1}
\ensuremath{\bigstar} } }

\newcommand{\joey}[1] { \textcolor{red} {
\ensuremath{\clubsuit} {\bf joey:}  {#1}
\ensuremath{\clubsuit} } }%\documentclass[manuscript]{aastex}

\begin{document}
%%%%%%%% TITLE, AUTHORS, PUBLICATION STATUS, and ABSTRACT %%%%%%%%%

\shorttitle{Fitting the undetected}
\shortauthors{}
\title{Fitting the undetected:  Light-curve feature estimation in the presence of carelessly reported non-detections}
\author{
Some wild \& crazy guys
}
%
%\altaffiltext{1}{Astronomy Department, University of California, Berkeley, CA, 94720-7450, USA}
%\altaffiltext{2}{Statistics Department, University of California, Berkeley, CA, 94720-7450, USA}
%\altaffiltext{*}{E-mail: {\tt jwrichar@stat.berkeley.edu}}

% Note the status (in preparation, MNRAS submitted date, ApJ accepted date, in press etc).
\slugcomment{in preparation}

\begin{abstract}
In multi-epoch imaging surveys, faint variable sources will not be
detectable at all epochs.  In principle, the best approach is to
perform photometric measurements anyway, so that there is a datum
required at at every epoch.  In practice, in most surveys we are
provided at some epochs only with upper limits; in some not even upper
limits but just the information that the star was not observed.  Here
we demonstrate with real data on periodic variables from the All-Sky
Automated Survey that these non-detections are nonetheless useful in
fitting and parameter estimation; their inclusion improves both
precision and accuracy.  One novel aspect of this work is that we do
not require the data source to include accurate---or even
any---information about the uncertainty variances on detections or the
values of the upper limits; we find that we can model these as latent
variables, even when they are different at every epoch.
\end{abstract}

%\keywords{stars: variables: general -- methods: data analysis -- methods: statistical -- techniques: photometric}

\section{Introduction}
\label{sec:intro}

\joey{What people actually do, what problems that could cause, and what people should do}

@@

\joey{RRL end at 120 kpc in Sesar et al. paper}

\joey{discuss in context of: parameter estimation (amplitude / period), discovery (we're gonna find a lot more variable sources if we take the non-detections into account), and classification; lossiness}

\joey{Numbers of LCs affected: Stripe 82, ASAS}

\section{Method}
\label{sec:method}

@@

Under the assumption that the inclusion (or not) of the source in the
catalog is based on (or very strongly related to) the measured flux
(rather than the true flux), the natural likelihood for the bit $q_i$,
on which the above likelihoods are conditioned is
\begin{eqnarray}\displaystyle
P(q_i|b_i,\sigma_i,\theta,I) &=& \left\{\begin{array}{ll}
  \displaystyle\int_{-\infty}^{b_i} N(f|\mu_i,s_\mu^2+\sigma_i^2)\,\dd f & \mbox{for $q_i=0$} \\
  \displaystyle\int_{b_i}^\infty N(f|\mu_i,s_\mu^2+\sigma_i^2)\,\dd f & \mbox{for $q_i=1$}
\end{array}\right.\label{eq:pofq}
\quad,
\end{eqnarray}
where the integrals end at an observed-flux limit $b_i$, which (as
mentioned above) is also a model parameter.  The motivation for this
expression is that it imagines all possible observations $f$ of the
star at time $t_i$ given the model and uncertainties, and finds the
fraction that would make the (imagined) detection cut.

@@

\subsection{Implementation}
\label{ss:implemenation}

\joey{Describe Python implementation here}

\section{Experiments}
\label{sec:experiments}

\subsection{Simulating Faint Mira Variables}
\label{ss:mirasim}

In this first experiment, we begin with a well-observed Mira variable from the ASAS Catalog of Variable Stars (ACVS, \citealt{acvs}) , ASAS 235627-4947.2.  This star has a pulsation period of 266.6286 days and a Lomb-Scargle amplitude of 2.38 mag (V band).  Note there are no non-detections in the ASAS light curve.

To simulate faint Mira stars from ASAS 235627-4947.2, we use the following procedure:
\begin{enumerate}
\item Convert the observed magnitudes, $m_i$, and errors, $s_{m,i}$ to fluxes, $f_i$ and flux errors, $s_{f,i}$ (we assume a V-band zero-point of $3.67\times10^{-9}$ erg/s/cm$^2$/\AA)
\item For a given flux dimming parameter, $d$, sample the new fluxes, $\tilde{f}_i$, from a Gaussian distribution centered around $f_i / d$ with standard deviation sampled from the empirical ASAS distribution of $s_{f,i} | f_i$. \joey{I found that relationship to be linear with a slope $\approx 1$, which is probably BS since the ASAS mag errors are all crap.  So I added a reference value to each flux error to ensure a mag limit $\approx$ 14.5 mag.}
\item Fluxes that are not at least 5$\sigma$ above zero are denoted as non-detections and their flux estimates (and errors) are censored.
\end{enumerate}
Folded light curves of ASAS 235627-4947.2, dimmed by four different values of $d$, are plotted in Figure \ref{fig:mirasim}.

\begin{figure*}
\begin{center}
\includegraphics[angle=0,width=6.5in]{../plots/mira_simulated.pdf}
\end{center}
\caption{ Folded light curves for data simulated from the Mira variable ASAS 235627-4947.2 using the prescription in \S\ref{ss:mirasim}.  Blue tick marks along the bottom axis denote phases where there were non-detections.  As the original flux measurements are dimmed by higher factors, the troughs of the sinusoidal light curve are censored, resulting in an incomplete view of the data.  \label{fig:mirasim}}
\end{figure*}

\subsection{Results of Fitting Miras}

\section{Results: ASAS Light Curves}
\label{sec:results}

We use the methodology of \cite{2011arXiv1106.2832R} to select the top ASAS Mira  and RR Lyrae, Fundamental Mode candidates.  Using a posterior probability threshold of 0.8 gives us 1720 Mira  and 1029 RR Lyrae candidates.

 \subsection{Mira Variables}

 Show P - A relationship before and after using the method

 \subsection{RR Lyrae Variables}

  Show P - A relationship before and after using the method

There is an RRL P - A relationship (strong linear anti-correlation)

\section{Discussion}
\label{sec:discussion}

What people can do, starting from the data-taking procedure.

limitations

\acknowledgements

\bibliography{non_detect}
